---
title: "Case Study 7"
author: "Piotr Suder"
date: "2023-04-15"
output: pdf_document
---

### Basic dependencies:

  - R>=4.1  # https://www.r-project.org/
  
  - RStudio>=1.4.1717  # https://posit.co/download/rstudio-desktop/
  
  
### R Package Dependencies

```{r}
options(warn=-1)

if (!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, magrittr, ggpubr, latex2exp, xtable, tidybayes, 
               survival, mice, ggsurvfit, gtsummary, survminer, dplyr, ggfortify)

dir.create("Figure", showWarnings = F)
theme_set(theme_classic(base_size = 12))
```

### System Information


```{r}
sessionInfo()
```

## Pre-processing

The following code is used for data preprocessing and imputation. Due to long runtime this part is commented out and the result stored in the data_full.RDS object is loaded.

```{r}
# data = read.csv('dukecathr.csv')
# 
# ID.vec = unique(data$RSUBJID)
# 
# death.adj = vector(length = nrow(data))
# time = vector(length = nrow(data))
# prev.cat = vector(length = nrow(data))
# 
# count = 0
# cutoff = 2*365
# for (i in 1:length(ID.vec)){
#   if (i %% 300 == 0)
#   {
#     cat('progress: ', (i/length(ID.vec))*100, '% \n')
#   }
#   id = ID.vec[i]
#   cath.dat = data[data$RSUBJID == id,]
#   tot.cath = nrow(cath.dat)
#   
#   #cat('id: ', id, ' tot.cath: ', tot.cath, '\n')
#   for (j in 1:tot.cath)
#   {
#     count = count + 1
#     if (j == 1)
#     {
#       prev.cat[count] = 0
#     }
#     if (j > 1)
#     {
#       t.diff = cath.dat$DAYS2LKA[j-1] - cath.dat$DAYS2LKA[j]
#       if (t.diff >= cutoff)
#       {
#         prev.cat[count] = 1
#       }
#       if (t.diff < cutoff)
#       {
#         prev.cat[count] = 2
#       }
#     }
#     if (j == tot.cath)
#     {
#       time[count] = cath.dat$DAYS2LKA[j]
#       death.adj[count] = cath.dat$DEATH[j]
#     }
#     if (j < tot.cath)
#     {
#       time[count] = cath.dat$DAYS2LKA[j] - cath.dat$DAYS2LKA[j+1]
#       death.adj[count] = 0
#     }
#   }
# }
# 
# data$prev_cat = prev.cat
# data$death_adj = death.adj
# data$survtime = time
# 
# # recode: DPCABG, DPMI, DPPCI, 
# # collapse: CHFSEV
# 
# # weight and height encode as BMI and maybe interact with age 
# # recde age to the median of each category
# 
# 
# # potentially exclude: CATHAPPR (only take left heart)
# 
# # TARGET: DAYS2LKA
# 
# # Make binary and set as did not happen if there is a catherization happening before it: DSCABG, DSMI, DSPCI,DSSTROKE,
# 
# 
# 
# data$past_CABG = 1*(!is.na(data$DPCABG) & data$DPCABG > 2*365) + 2*(!is.na(data$DPCABG) & data$DPCABG <= 2*365)
# data$past_MI = 1*(!is.na(data$DPMI) & data$DPMI > 365) + 2*(!is.na(data$DPMI) & data$DPMI <= 365)
# data$past_PCI = 1*(!is.na(data$DPPCI) & data$DPPCI > 365) + 2*(!is.na(data$DPPCI) & data$DPPCI <= 365)
# 
# 
# data$CHF_severity = 0*(data$CHFSEV == 0) + 1*(data$CHFSEV == 1 | data$CHFSEV == 2) + 2*(data$CHFSEV == 3 | data$CHFSEV == 4)
# data$CHF_severity[is.na(data$CHFSEV)] = NA
# 
# data$BMI = data$WEIGHT_R / ((data$HEIGHT_R/100)^2)
# 
# 
# data$age = 21*(data$AGE_G == 1) + 27*(data$AGE_G == 2) + 32*(data$AGE_G == 3) + 37*(data$AGE_G == 4) + 
#   42*(data$AGE_G == 5) + 47*(data$AGE_G == 6) + 52*(data$AGE_G == 7) + 57*(data$AGE_G == 8) + 
#   62*(data$AGE_G == 9) + 67*(data$AGE_G == 10) + 72*(data$AGE_G == 11) + 77*(data$AGE_G == 12) + 
#   85*(data$AGE_G == 13)
# 
# 
# # Change year into 3 categories: before coronary stents, before drug-eluting stents, and after both
# data = data|> mutate(year = case_when(
#   YRCATH_G %in% c(1,2) ~ 1, # coronary stents were approved by the FDA in 1994
#   YRCATH_G %in% c(3,4) ~ 2, # first drug-eluting stents were approved in 2003
#   TRUE ~ 3
# ))
# 
# # DSCABG, DSMI, DSPCI,DSSTROKE
# 
# data$subsequent_CABG = 1*(!is.na(data$DSCABG) & (data$DSCABG <= data$survtime))
# data$subsequent_MI = 1*(!is.na(data$DSMI) & (data$DSMI <= data$survtime))
# data$subsequent_PCI = 1*(!is.na(data$DSPCI) & (data$DSPCI <= data$survtime))
# data$subsequent_stroke = 1*(!is.na(data$DSSTROKE) & (data$DSSTROKE <= data$survtime))
# 
# ####################################
# 
# # Definitely exclude: FUPROTCL, RDAYSFROMINDEX, HXCHF (we already have CHF severity), NUMPRMI ?
# 
# data_proc = data %>% select(c('survtime', 'death_adj', 'age','year', 'GENDER', 'RACE_G', 'ACS', 'CHF_severity', 'past_CABG', 'past_MI', 'past_PCI', 
#                               'HXANGINA', 'HXCEREB', 'HXCOPD', 'HXDIAB', 'HXHTN', 'HXHYL', 'HXMI', 'HXSMOKE',
#                               'NUMPRMI', 'DIASBP_R', 'PULSE_R', 'SYSBP_R', 'CBRUITS', 'BMI', 'S3', 
#                               'CREATININE_R', 'HDL_R', 'LDL_R', 'TOTCHOL_R',
#                               'CATHAPPR', 'DIAGCATH', 'INTVCATH', 'CORDOM',
#                               'GRAFTST', 'LADST', 'LCXST', 'LMST', 'LVEF_R',
#                               'NUMDZV', 'PRXLADST', 'RCAST'))
# 
# data_complete = data_proc %>% na.omit()
# 
# head(data_complete)
# 
# nrow(data_complete)
# 
# data_mice = data_proc %>%
#   mutate(GENDER = factor(GENDER), RACE_G = factor(RACE_G), ACS = factor(ACS), year = factor(year),
#          CHF_severity = factor(CHF_severity), past_CABG = factor(past_CABG), past_MI = factor(past_MI), 
#          past_PCI = factor(past_PCI), HXANGINA = factor(HXANGINA), HXCEREB = factor(HXCEREB), HXCOPD= factor(HXCOPD),
#          HXDIAB = factor(HXDIAB), HXHTN = factor(HXHTN), HXHYL = factor(HXHYL), HXMI = factor(HXMI),
#          HXSMOKE = factor(HXSMOKE), CBRUITS = factor(CBRUITS), S3 = factor(S3), CATHAPPR = factor(CATHAPPR),
#          DIAGCATH = factor(DIAGCATH), INTVCATH = factor(INTVCATH), CORDO = factor(CORDOM))
# 
# 
# mice_obj <- mice(data = data_mice, m = 1)
# data_full <- complete(mice_obj)
# 
# saveRDS(data_full, "data_full.RDS")
```


```{r}
# Define colors
purple <- "#756bb1"
orange <- "#ffb50f"

data_full <- readRDS("data_full.RDS")
data.subject <- data_full %>% 
  mutate(DEATH2 = ifelse(death_adj == 0, "Censored", "Dead"),
         DAYS2LKA = survtime, DEATH = death_adj)

PlotKMCurve <- function(group_var, var_name = NULL) {
  if (is.null(var_name)) var_name <- group_var
  survfit2(Surv(DAYS2LKA + 0.1, DEATH) ~ get(group_var), data = data.subject) %>% 
    ggsurvfit() +
    add_confidence_interval() +
    labs(
      x = "Days",
      y = "Survival probability",
      fill = element_blank(), color = element_blank()
    )
}

fig1 <- PlotKMCurve("S3", "Third Heart Sound") +
  labs(subtitle = "Third Heart Sound") +
  scale_color_manual(values = c(purple, orange), labels = c("No (96%)", "Yes (4%)")) +
  scale_fill_manual(values = c(purple, orange), labels = c("No (96%)", "Yes (4%)"))
fig2 <- PlotKMCurve("past_CABG", "Past CABG") +
  labs(subtitle = "Closest Coronary Artery Bypass Surgery") +
  scale_color_manual(values = c("skyblue2", purple, orange), labels = c("Never (76%)", "More than 2 years ago (19%)",  "Within 2 years (5%)")) +
  scale_fill_manual(values = c("skyblue2", purple, orange), labels = c("Never (76%)", "More than 2 years ago (19%)",  "Within 2 years (5%)"))
fig3 <- PlotKMCurve("CATHAPPR", "CATHAPPR") +
  labs(subtitle = "Type of Cardiac Catheterization") +
  scale_color_manual(values = c(orange, "skyblue2", purple, "darkgrey"),
                     labels = c("Unknown (2%)", "Right (2%)",
                                "Left (85%)",
                                "Right and Left (11%)")) +
  scale_fill_manual(values = c(orange, "skyblue2", purple, "black"),
                    labels = c("Unknown (2%)", "Right (2%)",
                               "Left (85%)",
                               "Right and Left (11%)"))

pdf("Figure/eda1.pdf", height = 3.5, width = 6)
fig1
dev.off()
pdf("Figure/eda2.pdf", height = 3.5, width = 6)
fig2
dev.off()
pdf("Figure/eda3.pdf", height = 3.5, width = 6)
fig3
dev.off()
```

```{r}
plot(fig1)
```

```{r}
plot(fig2)
```

```{r}
plot(fig3)
```



Standardize using the method by Andrew Gelman.

```{r}
cont.names = c('age', 'DIASBP_R', 'PULSE_R', 'SYSBP_R', 'BMI', 'CREATININE_R', 'HDL_R', 'LDL_R', 'TOTCHOL_R',
               'GRAFTST', 'LADST', 'LCXST', 'LMST', 'LVEF_R', 'PRXLADST', 'RCAST')

data_full_std = data_full

for (i in 1:ncol(data_full_std))
{
  if (names(data_full_std)[i] %in% cont.names)
  {
    #print(names(data_full_std)[i] )
    data_full_std[,i] = (data_full_std[,i] - mean(data_full_std[,i]))/ (2*sd(data_full_std[,i]))
  }
}

head(data_full_std)

```

### AFT model

```{r}
survmodel = survreg(Surv(survtime + 0.1, death_adj)~1 + age + GENDER + RACE_G + year +  ACS + 
                      CHF_severity + past_CABG + past_MI + past_PCI +  
                      HXANGINA + HXCEREB + HXCOPD + HXDIAB + HXHTN +
                      HXHYL  + HXSMOKE + NUMPRMI + DIASBP_R + PULSE_R + SYSBP_R + #+ as.factor(HXMI)
                      + CBRUITS + BMI + S3 + CREATININE_R + HDL_R + LDL_R + TOTCHOL_R + CATHAPPR +
                      DIAGCATH + INTVCATH + CORDOM + GRAFTST + LADST + LCXST + LMST + LVEF_R +
                      NUMDZV + PRXLADST + RCAST, data = data_full_std, dist='weibull')
summary(survmodel)

xibeta = survmodel$linear.predictors


coeffs = survmodel$coefficients
lambda = exp(coeffs[1])
gamma = 1/survmodel$scale


low = confint(survmodel, level=.95)[-1,1]
high = confint(survmodel, level=.95)[-1,2]

df_results = data.frame(value = coeffs[-1], Q2.5= low, Q97.5 = high)
df_results$variable = rownames(df_results)

df_results$variable <- factor(df_results$variable, 
                              levels = df_results$variable[order(df_results$value)])
df_signif = df_results[df_results$Q2.5 > 0 | df_results$Q97.5 < 0,]
p0 = ggplot(df_signif, aes(y =variable, x = exp(value))) + 
  geom_errorbar(aes(xmax = exp(Q97.5), xmin =exp(Q2.5)),width=0.2) +
  geom_point(position = position_dodge(0.9)) +
  geom_vline(xintercept=1, color = "red", linetype = "dashed")+
  xlab(expression("decreased survival"  %<->% "increased survival")) + 
  ylab("Variable") + 
  labs(title = "95% CI for change in survival time, AFT model")+
  theme_bw()+
  theme( legend.position = "none", axis.title = element_text(size = 20), 
         plot.title = element_text(size = 22)) 



pdf("Figure/aft_fit.pdf", height = 7, width = 6)
p0
dev.off()

```

```{r}
plot(p0)
```


### Cox model

```{r}
coxmodel <- coxph(Surv(survtime + 0.1, death_adj) ~ age + GENDER + RACE_G + year +  ACS + 
                    CHF_severity + past_CABG + past_MI + past_PCI +  
                    HXANGINA + HXCEREB + HXCOPD + HXDIAB + HXHTN +
                    HXHYL  + HXSMOKE + NUMPRMI + DIASBP_R + PULSE_R + SYSBP_R + #+ as.factor(HXMI)
                    + CBRUITS + BMI + S3 + CREATININE_R + HDL_R + LDL_R + TOTCHOL_R + CATHAPPR +
                    DIAGCATH + INTVCATH + CORDOM + GRAFTST + LADST + LCXST + LMST + LVEF_R +
                    NUMDZV + PRXLADST + RCAST, data = data_full_std)
summary(coxmodel)

coeffs2 = coxmodel$coefficients


low2 = confint(coxmodel, level=.95)[,1]
high2 = confint(coxmodel, level=.95)[,2]

df_results_cox = data.frame(value = coeffs2, Q2.5= low2, Q97.5 = high2)
df_results_cox$variable = rownames(df_results_cox)

df_results_cox$variable <- factor(df_results_cox$variable, 
                                  levels = df_results_cox$variable[order(df_results_cox$value)])
df_signif_cox = df_results_cox[df_results_cox$Q2.5 > 0 | df_results_cox$Q97.5 < 0,]

p1 = ggplot(df_signif_cox, aes(y =variable, x = exp(value))) + 
  geom_errorbar(aes(xmax = exp(Q97.5), xmin = exp(Q2.5)),width=0.2) +
  geom_point(position = position_dodge(0.9)) +
  geom_vline(xintercept=1, color = "red", linetype = "dashed")+
  xlab(expression("lower hazard"  %<->% "higher hazard")) +
  ylab("Variable") + 
  labs(title = "95% CI hazard ratio, Cox model")+
  theme_bw()+
  theme( legend.position = "none", axis.title = element_text(size = 20), 
         plot.title = element_text(size = 22)) 


pdf("Figure/cox_fit.pdf", height = 7, width = 6)
p1
dev.off()

#### Table
sigma = survmodel$scale
##### S3:


exp(-coeffs["S31"]/sigma)
exp(coeffs2["S31"])

### Capp right
exp(-coeffs["CATHAPPR1"]/sigma)
exp(coeffs2["CATHAPPR1"])

### Capp left
exp(-coeffs["CATHAPPR2"]/sigma)
exp(coeffs2["CATHAPPR2"])

### Capp left and right
exp(-coeffs["CATHAPPR3"]/sigma)
exp(coeffs2["CATHAPPR3"])



### Past CABG1
exp(-coeffs["past_CABG1"]/sigma)
exp(coeffs2["past_CABG1"])

### Past CABG1
exp(-coeffs["past_CABG2"]/sigma)
exp(coeffs2["past_CABG2"])

```

```{r}
plot(p1)
```

## Diagnostics

```{r}
source("http://myweb.uiowa.edu/pbreheny/7210/f18/notes/fun.R")

sfit <- survfit(coxmodel)
H0 <- -log(sfit$surv)
H <- approxfun(c(0, sfit$time), c(0, H0), method='constant')
e1 <- H(coxmodel$y[,1])*exp(coxmodel$linear.predictors)
e2 <- coxmodel$y[,2]-residuals(coxmodel)
head(e1)
head(e2)

# Slide 5: Diagnostic plot
efit <- survfit(Surv(e1, coxmodel$y[,2])~1)
lim <- c(0,15)

pdf("Figure/cox_snail_resid.pdf", height = 3.5, width = 6)
plot(efit, fun='cumhaz', mark.time=FALSE, bty='n', conf.int=FALSE, lwd=1, las=1,
     xlab='Residual', ylab='Cumulative hazard', xlim=lim, ylim=lim)
ciband(efit, fun=function(x) -log(x))
lines(lim, lim, col='red', lwd=1)
dev.off()

```

```{r}
plot(efit, fun='cumhaz', mark.time=FALSE, bty='n', conf.int=FALSE, lwd=1, las=1,
     xlab='Residual', ylab='Cumulative hazard', xlim=lim, ylim=lim)
ciband(efit, fun=function(x) -log(x))
lines(lim, lim, col='red', lwd=1)
```


### Oout-of sample-performance - Concordance

```{r}
get.concordance = function(pred_test, truth_test, death)
{
  nvalid = length(pred_test)
  agree.count = 0
  pair.count = 0
  for (i in 2:nvalid)
  {
    for (j in 1:(i-1))
    {
      pair.count = pair.count + death[j]
      agree.count = agree.count + death[j]*((pred_test[i] >= pred_test[j]) == (truth_test[i] >= truth_test[j]))
    }
  }
  
  concord = agree.count / pair.count
  return(concord)
}

ndata = nrow(data_full_std)
death_ind = which(data_full$death_adj == 1)
nvalid = floor(0.1*nrow(data_full_std))

set.seed(4)
index = sample(1:nrow(data_full_std), size = nvalid, replace = FALSE)

test_data = data_full_std[index,]
train_data = data_full_std[-index,]

# AFT

survtrain_aft = survreg(Surv(survtime + 0.1, death_adj)~1 + age + GENDER + RACE_G + year +  ACS + 
                          CHF_severity + past_CABG + past_MI + past_PCI +  
                          HXANGINA + HXCEREB + HXCOPD + HXDIAB + HXHTN +
                          HXHYL  + HXSMOKE + NUMPRMI + DIASBP_R + PULSE_R + SYSBP_R + #+ as.factor(HXMI)
                          + CBRUITS + BMI + S3 + CREATININE_R + HDL_R + LDL_R + TOTCHOL_R + CATHAPPR +
                          DIAGCATH + INTVCATH + CORDOM + GRAFTST + LADST + LCXST + LMST + LVEF_R +
                          NUMDZV + PRXLADST + RCAST, data = train_data, dist='weibull')

pred_test_aft <- predict(survtrain_aft, newdata=test_data, type='response', se=FALSE)
truth_test = test_data$survtime

coxmodel_train <- coxph(Surv(survtime + 0.1, death_adj) ~ age + GENDER + RACE_G + year +  ACS + 
                          CHF_severity + past_CABG + past_MI + past_PCI +  
                          HXANGINA + HXCEREB + HXCOPD + HXDIAB + HXHTN +
                          HXHYL  + HXSMOKE + NUMPRMI + DIASBP_R + PULSE_R + SYSBP_R + #+ as.factor(HXMI)
                          + CBRUITS + BMI + S3 + CREATININE_R + HDL_R + LDL_R + TOTCHOL_R + CATHAPPR +
                          DIAGCATH + INTVCATH + CORDOM + GRAFTST + LADST + LCXST + LMST + LVEF_R +
                          NUMDZV + PRXLADST + RCAST, data = train_data)

pred_test_cox <- predict(coxmodel_train, newdata=test_data, type='risk', se=FALSE)


survtrain_aft_loglog = survreg(Surv(survtime + 0.1, death_adj)~1 + age + GENDER + RACE_G + year +  ACS + 
                                 CHF_severity + past_CABG + past_MI + past_PCI +  
                                 HXANGINA + HXCEREB + HXCOPD + HXDIAB + HXHTN +
                                 HXHYL  + HXSMOKE + NUMPRMI + DIASBP_R + PULSE_R + SYSBP_R + #+ as.factor(HXMI)
                                 + CBRUITS + BMI + S3 + CREATININE_R + HDL_R + LDL_R + TOTCHOL_R + CATHAPPR +
                                 DIAGCATH + INTVCATH + CORDOM + GRAFTST + LADST + LCXST + LMST + LVEF_R +
                                 NUMDZV + PRXLADST + RCAST, data = train_data, dist='lognormal')


pred_test_aft_loglog <- predict(survtrain_aft_loglog, newdata=test_data, type='response', se=FALSE)


# AFT Concordance
concord.aft = get.concordance(pred_test = pred_test_aft, truth_test = truth_test, death = data_full_std$death_adj)
print(concord.aft)

# Cox Concordance - the prediction is for risk, so we take negative
concord.cox = get.concordance(pred_test = -pred_test_cox, truth_test = truth_test, death = data_full_std$death_adj)
print(concord.cox)
```